\chapter{METHODS}
\label{chap:methods}
This chapter describes the computational workflow used to retrieve, process and utilise omics data to classify established \gls{lusc} molecular subtypes and to interpret the molecular features driving classification.

% TODO: add a cohort assembly tikz figure

\section{Data Acquisition and Label Matching}
\label{subsec:data_acquisition}

Matched omics data for the \gls{tcga}-\gls{lusc} cohort were retrieved through the UCSC Xena Platform \parencite{Goldman2020}, which hosts processed \gls{tcga} data harmonised from original \gls{tcga} releases.

\section{Omics Preprocessing}
\label{subsec:omics_preprocessing}

Preprocessing was performed separately for each modality to preserve the native scale and biological specific to respective platforms.

\subsection{Gene Expression}
RSEM-normalised read counts were provided in $\log_2(x+1)$-transformed form and required no additional normalisation. Genes with median $\log_2$ expression below 1.0 across all samples were removed as unexpressed. The 2,000 most variable genes were then selected by \gls{mad} filtering.

\subsection{\gls{dna} Methylation}
Level 3 beta values, representing the ratio of methylated to total probe intensity and ranging from 0 to 1, were used directly. CpG probes missing in more than 20\% of samples were removed, and remaining missing values were imputed using the per-probe median. Probes with standard deviation below 0.05 were removed as near-invariant, and the top 2,000 most variable probes were retained by \gls{mad}. For downstream gene-level interpretation, CpG probe IDs were mapped to gene symbols using Illumina HumanMethylation450 annotation; probes beginning with ``rs'' were treated as SNP probes and excluded from methylation-specific biological interpretation.

\subsection{\gls{mirna} Expression}
$\log_2(\mathrm{RPM}+1)$-transformed \gls{mirna} expression values were used without further normalisation. Missing values in this data type indicate zero reads (RPM = 0), and were therefore filled with zero rather than imputed. \gls{mirna}s missing in more than 50\% of samples were removed, followed by low-expression filtering (median $\log_2(\mathrm{RPM}+1) < 1.0$) and \gls{mad} filtering to retain 200 most variable mature \gls{mirna}s.

\subsection{Copy Number Variation}
GISTIC2-thresholded discrete copy number values (integer scale: $-2, -1, 0, 1, 2$) were used directly. No normalisation was applied because the data encode categorical copy number states. Zero-variance genes were checked. The top 2,000 most variable genes were retained by \gls{mad}.

\subsection{RPPA}
Normalised RPPA protein abundance values were used directly. Proteins missing in more than 20\% of samples were removed, and remaining missing values were imputed using the per-protein median. No additional \gls{mad} filtering was applied because the post-imputation feature count was already appropriate for the available sample size.

\section{Dataset Separability}
Principal component analysis (PCA) was performed on each omics modality separately to characterise the geometric separability of the subtypes within each feature space. StandardScaler normalisatin was applied to each modality, and minimum number of principal components explaining \geq 80\% of cumulative variance was determined. 

% TODO: cite scipy.stats library here, as i used squareform for euclidean distance thing.. other methods too.
% TODO: cite Rousseeuw, 1987 for silhouette score.  
% TODO: cite (Anderson, 2001) for permanova.  
Separability of subtypes was quantified using two complementary measures. The silhouette score was computed on the 80\%-variance PCA subspace, providing a sample-level measure of cluster cohesion and separation ranging from -1 (misassigned) to +1 (well-separated). Per-subtype mean silhouette scores were computed to identify subtypes with differential cluster membership. PERMANOVA with 999 permutations was applied to the same PCA subspace to test the null hypothesis that subtype group centroids are statistically indistinguishable from permuted label assignments; (pseudo-F statistic; permutation p-value). PCA projections were visualised on PC1 and PC2 with 95\% confidence ellipses (n_std = 2.0) and subtype centroids.